% ============================================================
% Paper 5: IEEE / OFC Conference Style
% "Photonic Compilation of MedGemma-4B:
%  A Complete SVD-Clements Pipeline for Medical AI Inference"
% IEEE Photonics Technology Letters / OFC 2026 format
% ============================================================
\documentclass[journal]{IEEEtran}

\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{cite}
\usepackage{url}
\usepackage{xcolor}
\usepackage{multirow}

\hypersetup{colorlinks=true, linkcolor=blue!70!black, citecolor=blue!70!black}

% ----------------------------------------------------------------
\title{Photonic Compilation of MedGemma-4B:\\
       A Complete SVD-Clements Pipeline for\\
       Medical AI Inference on Silicon Photonic Chips}

\author{
  Linda~H.
  \thanks{Independent Research. E-mail: photomedgemma@github.com}
}

% ----------------------------------------------------------------
\begin{document}
\maketitle

% ----------------------------------------------------------------
\begin{abstract}
We present a complete software-to-photonics compilation pipeline that maps
the weight matrices of MedGemma-4B-IT --- a 4\,B-parameter multimodal
medical vision-language model --- to phase settings for silicon photonic
Mach-Zehnder interferometer (MZI) meshes.
The pipeline uses singular value decomposition (SVD) to factor each weight
matrix into unitary components, followed by the Clements algorithm to
decompose each unitary into cascaded 2$\times$2 MZI rotations.
Validation on real MedGemma activations from five clinical breast cancer
images demonstrates compilation fidelity $\varepsilon < 8\times10^{-15}$
(machine precision) for all four layer-0 attention projections.
A full rank sweep from $r=64$ to full rank characterises the accuracy-hardware
trade-off at each rank.
The photonic architecture targets 220\,nm SOI and projects $\sim37\times$
energy efficiency improvement over GPU-based inference, enabling deployment in
resource-constrained clinical environments.
\end{abstract}

\begin{IEEEkeywords}
Silicon photonics, optical neural networks, MZI mesh, Clements decomposition,
singular value decomposition, medical AI, low-resource deployment.
\end{IEEEkeywords}

% ----------------------------------------------------------------
\section{Introduction}
\label{sec:intro}

\IEEEPARstart{P}{hotonic} neural network accelerators exploit the speed of
light and the passive nature of optical interference to perform matrix-vector
multiplication with energy efficiency approaching the thermodynamic limit
\cite{shen2017}.
For inference from fixed, trained models, the weight matrices are constant ---
enabling \emph{static compilation}: the weights are encoded once into the
phase settings of an MZI mesh and need not be updated during inference.
This eliminates the dominant energy cost of electronic matrix multipliers
(weight memory reads and digital multiply-accumulate operations).

Google's MedGemma-4B-IT model \cite{medgemma2024} achieves
clinical-grade accuracy on medical imaging and diagnostic reasoning tasks,
but requires GPU infrastructure ($\sim300$\,W) incompatible with
energy-constrained hospital deployments.
The specific context motivating this work is Kenyatta National Hospital (KNH)
in Nairobi, Kenya --- the largest referral hospital in East and Central Africa
--- where data sovereignty regulations prohibit cloud inference, and local
power infrastructure cannot support continuous GPU operation.

We present PhotoMedGemma: an open-source compiler and chip architecture that
translates MedGemma's attention layer weight matrices to 220\,nm SOI photonic
phase maps, with validation at machine precision.

% ----------------------------------------------------------------
\section{Compiler Architecture}
\label{sec:compiler}

\subsection{SVD Decomposition}

Each weight matrix $W \in \mathbb{R}^{m\times n}$ is decomposed via
truncated SVD:
\begin{equation}
  W \approx W_r = U_r \Sigma_r V_r^\dagger,
  \label{eq:svd}
\end{equation}
where $U_r \in \mathbb{C}^{m\times r}$ (partial unitary),
$\Sigma_r = \mathrm{diag}(\sigma_1,\ldots,\sigma_r)$,
$V_r^\dagger \in \mathbb{C}^{r\times n}$ (partial unitary).
Each partial unitary is embedded in a full $\max(m,n)\times\max(m,n)$ unitary
by extending with an orthonormal complement.

\subsection{Clements Decomposition}

Any $N\times N$ unitary $U$ is factored as \cite{clements2016}:
\begin{equation}
  U = D \cdot T_{N(N-1)/2} \cdots T_2 \cdot T_1,
  \label{eq:clements}
\end{equation}
where $D = \mathrm{diag}(e^{i\psi_j})$ and each $T_k$ acts on modes
$(m_k, m_k{+}1)$ as:
\begin{equation}
  T(\theta,\varphi) =
  \begin{pmatrix}
    \cos\tfrac{\theta}{2} & -e^{-i\varphi}\sin\tfrac{\theta}{2} \\
    e^{i\varphi}\sin\tfrac{\theta}{2} & \cos\tfrac{\theta}{2}
  \end{pmatrix}.
  \label{eq:T}
\end{equation}

The MZI nulling angles for mode pair $(r-1, r)$ eliminating element
$U_{r,c}$ are:
\begin{equation}
  \varphi = \angle U_{r,c} - \angle U_{r-1,c} + \pi,
  \quad
  \theta = 2\arctan\!\left(\tfrac{|U_{r,c}|}{|U_{r-1,c}|}\right).
  \label{eq:nulling}
\end{equation}

Forward-pass application proceeds in \emph{descending column order},
followed by multiplication by the diagonal phase screen $D$.

\subsection{Phase Quantisation}

Phases are quantised to $B=12$-bit DAC codes.
The worst-case phase error is $\Delta\phi_{\max} = \pi/2^{12} \approx 0.7$\,mrad,
producing matrix reconstruction error $\varepsilon_{\rm quant} \approx
\pi N / (\sqrt{3}\cdot2^B) \approx 0.05$ for $N=64$ --- negligible for
medical AI inference.

% ----------------------------------------------------------------
\section{Silicon Photonic Platform}
\label{sec:platform}

\begin{table}[h]
\centering
\caption{220\,nm SOI platform parameters (imec iSiPP50G / AIM Photonics)}
\label{tab:platform}
\begin{tabular}{ll}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Waveguide core & Si, 450\,nm $\times$ 220\,nm \\
Cladding & SiO$_2$ \\
Wavelength & 1310\,nm (O-band) \\
Phase shifter & TiN heater, $P_\pi = 10$--$20$\,mW \\
Thermo-optic coeff. & $dn/dT = 1.84\times10^{-4}$\,K$^{-1}$ \\
Propagation loss & 2--3\,dB/cm \\
Coupling loss (grating) & 2--4\,dB/coupler \\
Detector & Ge PD, $>$40\,GHz, 0.9\,A/W \\
Single-chip size & 10\,mm $\times$ 10\,mm \\
MZIs per chip (rank-64) & 4,096 \\
\bottomrule
\end{tabular}
\end{table}

Each chip implements a single $64\times64$ Clements unitary mesh
(2,016 MZIs in the rank-64 case) in an 8\,mm $\times$ 8\,mm active area,
with 64 grating couplers per edge for fiber array coupling.
Phase control uses 4,096 heater channels with 16-bit DAC on SPI bus.

% ----------------------------------------------------------------
\section{Experimental Validation}
\label{sec:experiments}

\subsection{Setup}

MedGemma-4B-IT was run on five breast cancer PNG images
on a Kaggle T4 GPU (4-bit NF4 quantisation).
For each image, the layer-0 hidden state inputs and attention projection
outputs were captured via forward hooks, along with the dequantised weight
matrices $W_q, W_k, W_v, W_o$.

\subsection{Compilation Fidelity}

Table~\ref{tab:fidelity} reports compilation fidelity for all four
layer-0 attention projections.
Chip fidelity is at machine precision ($\approx 10^{-15}$) in all cases,
confirming that the Clements algorithm faithfully encodes the target unitary.
The 4-bit NF4 weight extraction error is $< 0.22\%$ --- confirming
that the compiled weights accurately represent the true MedGemma weights.

\begin{table}[h]
\centering
\caption{Compilation fidelity for MedGemma layer-0 attention projections.}
\label{tab:fidelity}
\begin{tabular}{lcccc}
\toprule
\textbf{Proj.} & \textbf{Shape} & $\varepsilon_U$ & $\varepsilon_{V^\dagger}$ & $\varepsilon_{\rm quant}$ \\
\midrule
q\_proj & $2048\!\times\!2560$ & $7.0\!\times\!10^{-15}$ & $7.9\!\times\!10^{-15}$ & 0.0017 \\
k\_proj & $1024\!\times\!2560$ & $5.2\!\times\!10^{-15}$ & $7.9\!\times\!10^{-15}$ & 0.0019 \\
v\_proj & $1024\!\times\!2560$ & $5.2\!\times\!10^{-15}$ & $8.0\!\times\!10^{-15}$ & 0.0022 \\
o\_proj & $2560\!\times\!2048$ & $7.8\!\times\!10^{-15}$ & $6.9\!\times\!10^{-15}$ & 0.0020 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Rank-Accuracy-Hardware Trade-off}

Table~\ref{tab:rank_sweep} shows the SVD approximation error, MZI count, and
chip area for q\_proj at each rank.
The chip computes the rank-$r$ approximation to machine precision at all
ranks; the reported error is the inherent SVD approximation error.

\begin{table}[h]
\centering
\caption{Rank sweep for q\_proj ($2048\times2560$, full rank = 2048).
         Chip fidelity $\approx10^{-15}$ at all ranks.}
\label{tab:rank_sweep}
\begin{tabular}{rrcrc}
\toprule
\textbf{Rank} & \textbf{Energy} & $\varepsilon_{\rm SVD}$ &
\textbf{MZIs/mesh} & \textbf{Area (cmp)} \\
\midrule
    64 & 28.8\% & 0.186 &        2,016 &   1.6\,mm$^2$ \\
   128 & 38.2\% & 0.166 &        8,128 &   6.6\,mm$^2$ \\
   256 & 52.8\% & 0.142 &       32,640 &  26.2\,mm$^2$ \\
   512 & 72.6\% & 0.108 &      130,816 & 104.9\,mm$^2$ \\
  1024 & 92.3\% & 0.061 &      523,776 & 419.4\,mm$^2$ \\
  2048 & 100\%  & 0.000 &    2,096,128 & 1677.7\,mm$^2$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent
Compact Si: 80\,\textmu m stage $\times$ 5\,\textmu m pitch per MZI.

% ----------------------------------------------------------------
\section{System Architecture and Power Budget}
\label{sec:system}

\subsection{Multi-Chip Module}

A rank-64 layer-0 attention block requires 8 chips (4 projections $\times$
2 meshes/projection).
With 8-wavelength WDM parallelism for 8 attention heads, this reduces to
1 chip per projection = \textbf{4 chips} for the full layer-0 attention block.

For the full MedGemma-4B language model (46 layers, 7 weight matrices/layer)
with WDM:
\begin{equation}
  N_{\rm chips} = \frac{46 \times 7 \times 2}{8} \approx 81 \text{ chips.}
\end{equation}

\subsection{Power Budget}

\begin{table}[h]
\centering
\caption{System power budget (full inference, rank-64, 46 layers).}
\label{tab:power}
\begin{tabular}{lrl}
\toprule
\textbf{Component} & \textbf{Power} & \textbf{Notes} \\
\midrule
Laser (8$\times$ WDM)    & 800\,mW & 100\,mW each \\
Phase shifters (static)  &   0\,mW & Steady-state \\
Photodetectors           &  50\,mW & Ge PD array \\
FPGA (softmax, LN)       &   5\,W  & \\
Memory (KV cache)        &   2\,W  & LPDDR5 \\
\midrule
\textbf{Total}           & $\approx$\textbf{8\,W} &
   \textbf{vs.\ $\sim$300\,W (A100)} \\
\bottomrule
\end{tabular}
\end{table}

At 1 token/ms: \textbf{8\,mJ/token} vs.\ 300\,mJ/token for A100 GPU ---
a $37\times$ improvement in energy per token.

% ----------------------------------------------------------------
\section{Discussion}
\label{sec:discussion}

\paragraph{Compilation correctness.}
The critical implementation detail is the Clements T-matrix convention
(equation~\ref{eq:T}) combined with descending column application order.
Incorrect ordering (ascending) or incorrect T-matrix orientation produces
errors $\sim 1$--$2$ (larger than the signal), not $\sim 10^{-15}$.
The correct convention was verified by unit-testing the compiled mesh against
the target unitary; both $\varepsilon_U$ and $\varepsilon_{V^\dagger}$ must be
at machine precision before a phase map is accepted.

\paragraph{SVD approximation vs.\ chip fidelity.}
These are orthogonal quantities.
Chip fidelity measures the compilation algorithm's correctness (always
$\approx 10^{-15}$).
SVD approximation error measures how much the rank-$r$ truncation deviates
from the full weight (a design trade-off, ranging from 0 at full rank to 0.67
at rank 64 for v\_proj).

\paragraph{Path to fabrication.}
The compiled phase map (\texttt{phase\_map.json}) and GDS layout files
(\texttt{output/real/gds/}) are compatible with 220\,nm SOI PDKs from imec,
AIM Photonics, and GlobalFoundries.
Tape-out requires a multi-project wafer (MPW) submission, available through
the European Silicon Photonics Alliance (ePIXfab) or AIM Photonics open
foundry access.

% ----------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

We have presented a complete SVD-Clements compilation pipeline for
MedGemma-4B-IT, validated at machine precision on real clinical images.
The PhotoMedGemma framework produces ready-to-fabricate GDS layouts for
220\,nm SOI silicon photonics, with a full rank sweep characterising the
accuracy-hardware trade-off from rank 64 to 2048.
The projected $37\times$ energy efficiency improvement over A100 GPU inference
makes this architecture practical for deployment at KNH Nairobi and equivalent
settings throughout the global south.
All code, compiled phase maps, and experimental data are open-source.

% ----------------------------------------------------------------
\bibliographystyle{IEEEtran}
\begin{thebibliography}{9}

\bibitem{shen2017}
Y.\ Shen et al., ``Deep learning with coherent nanophotonic circuits,''
\textit{Nature Photonics}, vol.\ 11, no.\ 7, pp.\ 441--446, Jul.\ 2017.

\bibitem{medgemma2024}
Google DeepMind, ``MedGemma Technical Report,'' 2024.

\bibitem{clements2016}
W.R.\ Clements, P.C.\ Humphreys, B.J.\ Metcalf, W.S.\ Kolthammer, and I.A.\ Walmsley,
``Optimal design for universal multiport interferometers,''
\textit{Optica}, vol.\ 3, no.\ 12, pp.\ 1460--1465, Dec.\ 2016.

\bibitem{reck1994}
M.\ Reck, A.\ Zeilinger, H.J.\ Bernstein, and P.\ Bertani,
``Experimental realization of any discrete unitary operator,''
\textit{Phys.\ Rev.\ Lett.}, vol.\ 73, no.\ 1, p.\ 58, Jul.\ 1994.

\bibitem{hamerly2019}
R.\ Hamerly, L.\ Bernstein, A.\ Sludds, M.\ Solja\v{c}i\'{c}, and D.\ Englund,
``Large-scale optical neural networks based on photoelectric multiplication,''
\textit{Phys.\ Rev.\ X}, vol.\ 9, p.\ 021032, May 2019.

\bibitem{pai2023}
S.\ Pai et al., ``Experimentally realized in situ backpropagation for deep
learning in photonic neural networks,'' \textit{Science}, vol.\ 380, no.\ 6643,
pp.\ 398--404, Apr.\ 2023.

\bibitem{bandyopadhyay2022}
S.\ Bandyopadhyay et al., ``Single chip photonic deep neural network with
accelerated training,'' \textit{arXiv:2208.01623}, Aug.\ 2022.

\bibitem{gemma3_2024}
Google DeepMind, ``Gemma 3 Technical Report,'' \textit{arXiv:2408.00118}, 2024.

\end{thebibliography}

\end{document}
